---
title: "Appendix"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


<!-- EDA: -->
<!-- - Qualitative Variable Analysis - distribution(pie chat), side-by-side boxplot on response variable -->
<!-- - Quantitative Variables - distribution(histograms), transformations as appropriate, pairwise scatter plots, correlation matrix, multicollinearity -->
<!-- - Transform response variable -->
<!-- - Consolidate M,F for sex due to identical distributions; further explore the M,F distributions to confirm with plots -->

<!-- Modeling: -->
<!-- - First Order -->
<!-- - First Order: remove whole weight -->
<!-- - First Order: remove diameter -->
<!-- - Center predictors before polynomial regression -->
<!-- - Second Order -->
<!-- - Age has: Linear relationship length, diameter, height; Quadratic relationship with whole, shucked, viscera, shell -->
<!-- - Second Order: because of residuals vs fitted plot no longer looking slightly quadratic, we decide to continue further with the second order model -->
<!-- - Second Order with removal of excess predictors with significant multicollinearity -->
<!-- - Stepwise AIC,BIC -->
<!-- - forward, backward, bidirectional agree on final models -->
<!-- - length is not in the final model under bic criteria -->
<!-- - Best subset Regression -->
<!-- - Model with 11 predictor variables is best, without viscera^2 -->

<!-- Model Validation -->
<!-- - AIC -->
<!-- - BIC -->
<!-- - Cp? -->
<!-- - Press? -->
<!-- - root mean squared prediction error (RMSPE) -->

<!-- Candidate Models -->
<!-- - -->

<!-- Conclusion -->
<!-- - -->


# A.1 Loading Data

```{r}
#Read data, change col names, reorganize columns, set factors for qualitative variables
abalone <- read.csv("C:/Users/rly75/OneDrive/Desktop/206/abalone.txt", header=FALSE)
abalone = setNames(abalone,c('sex','length','diameter','height','whole','shucked','viscera','shell','rings'))

abalone$sex = as.factor(abalone$sex)
```

# A.2 Distribution of Data

```{r}
#Pie chart for qualitative variables, histogram for quantitative variables
n <- nrow(abalone)
lbls <- c('Female','Male','Infant')
pct <- round(100*table(abalone$sex)/n)
lab <- paste(lbls,pct)
lab <- paste(lab,'%',sep='')
#lab

par(mfrow=c(1,1))
pie(table(abalone$sex),labels=lab,col=c('pink','blue','green'),main='Abalone Sex')
hist(abalone$rings)

par(mfrow=c(1,1))
hist(abalone$length)
hist(abalone$diameter)
hist(abalone$height)
hist(abalone$whole)
hist(abalone$shucked)
hist(abalone$viscera)
hist(abalone$shell)

boxplot(abalone$rings~abalone$sex,main='Abalone: side-by-side box plot by sex',
xlab='Sex',ylab='Rings',col=rainbow(4))
```

# A.3 Initial First Order Fit 

```{r}
#try a first order fit with no transformations
fit = lm(rings~sex+length+diameter+height+whole+shucked+viscera+shell,data=abalone)
summary(fit)
plot(fit)

panel.cor <- function(x, y) {
    # usr <- par('usr') on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- round(cor(x, y, use = "complete.obs"), 2)
    txt <- paste0("R = ", r)
    cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}

pairs(~rings+length+diameter+height+whole+shucked+viscera+shell,data=abalone,lower.panel=panel.cor)
```

# A.4 Linear Relationships Between Variables

```{r}
#Linear relationships between response and the following predictors
plot(abalone$length,abalone$rings)
plot(abalone$diameter,abalone$rings)
plot(abalone$height,abalone$rings)
```

# A.5 Quadratic Relationships Between Variables

```{r}
#quadratic relationship between response and the following predictors
plot(abalone$whole,abalone$rings)
plot(abalone$shucked,abalone$rings)
plot(abalone$viscera,abalone$rings)
plot(abalone$shell,abalone$rings)
```

# A.6 Combining Male and Female Classes

```{r}
#Consider consolidating sex into two classes: infant(I) or not infant(NI)
abalone['infant'] = ifelse(abalone$sex=='I','I','NI')
abalone$infant = as.factor(abalone$infant)

boxplot(abalone$rings~abalone$infant,main='Abalone: side-by-side box plot by Age Status',
xlab='Sex',ylab='Age Status',col=rainbow(4))
```


# A.7 Boxcox Procedure and Transformation of Response Variable

```{r}
#check to see if any transformation is appropriate for response variable, lambda = 0 -> log transform
library(MASS)
bc = boxcox(fit)
lambda = bc$x[which.max(bc$y)]
lambda
```


```{r}
#perform log transform on response variable
abalone$rings = log(abalone$rings)
```


# A.8 Ring Distribution After Transformation

```{r}
#Check distributions post transform
hist(abalone$rings)

boxplot(abalone$rings~abalone$sex,main='Abalone: side-by-side box plot by sex',
xlab='Sex',ylab='Age',col=rainbow(3))

boxplot(abalone$rings~abalone$infant,main='Abalone: side-by-side box plot by Age Status',
xlab='Sex',ylab='Age Status',col=c('green','red'))
```

# A.9 Splitting the Data Set into Traning (70%) and Testing (30%) Sets 

```{r}
#split data into training and validating subsets
set.seed(100)
ind <- sample(nrow(abalone), 2924, replace = FALSE) #70% train, 30% valid
train <- abalone[ind, ] #training set
valid <- abalone[-ind, ] #validation/test set
```

# A.10 First Order Model with Transformed Response

```{r}
#Fit a first order model on training set (with transformed response variable)
fit1 = lm(rings~.-infant,data=train)
summary(fit1)
plot(fit1)
```



# A.11 Multicollinearity (1)

```{r}
#significant multicollinearity for all quantitative predictors outside of height
#most notably whole weight, likely because it is a function of shucked,viscera,shell
#length, diameter too because of high correlation
vif = diag(solve(cor(abalone[,c("length","diameter","height","whole","shucked","viscera","shell")])))
vif
```
# A.12 Coefficient of Partial Determination and Added Variable Plots for Whole Weight

```{r}
#Look at coefficient of partial determination by regressing response onto predictors, with and then without what you are testing on
#Look at added variable plots to see if we should retain or remove predictors
fit_whole = lm(whole~sex+length+diameter+height+shucked+viscera+shell,data=train)
fit_remove_whole = lm(rings~sex+length+diameter+height+shucked+viscera+shell,data=train)

#SSR(whole given all) = 2.301   
#SSE(.-whole) = 0.3488 + 122.585
#Coefficient of partial determination = 2.301 / 2.301 + 122.585 = 0.0184248
#So whole weight is of little additional help in explaining age when the remaining predictors are already in the model 
fit_whole_last = lm(rings~sex+length+diameter+height+shucked+viscera+shell+whole,data=train) 
anova(fit_whole_last)

plot(fit_whole$residuals,fit_remove_whole$residuals,main='Added Variable Plot for Whole Weight')
```

# A.13 Variance Inflation Factors After Removal of Whole Weight

```{r}
#Relook at VIFs after removing whole weight
vif = diag(solve(cor(abalone[,c("length","diameter","height","shucked","viscera","shell")])))
vif
```

# A.14 Coefficient of Partial Determination and Added Variable Plot for Diameter

```{r}
#Look at coefficient of partial determination by regressing response onto predictors, with and then without what you are testing on
#Look at added variable plots to see if we should retain or remove predictors
fit_diameter = lm(diameter~sex+length+height+shucked+viscera+shell,data=train)
fit_remove_diameter = lm(rings~sex+length+height+shucked+viscera+shell,data=train)

#SSR(whole given all) = 1.560   
#SSE(.-whole) = 1.560    + 124.886     
#Coefficient of partial determination = 1.560 / 1.560 + 124.886      = 0.01233728
#So whole weight is of little additional help in explaining age when the remaining predictors are already in the model 
fit_diameter_last = lm(rings~sex+length+height+shucked+viscera+shell+diameter,data=train) 
anova(fit_diameter_last)

plot(fit_diameter$residuals,fit_remove_diameter$residuals,main='Added Variable Plot for Diameter')
```


# A.15 Variance Inflation Factors after Removing Diameter

```{r}
#Relook at VIFs after removing diameter, conclude the first order model without whole,diameter can be further investigated
vif = diag(solve(cor(abalone[,c("length","height","shucked","viscera","shell")])))
vif
```
# A.16 Reduced First Order Model

```{r}
#Examine plots for reduced first order model
fit1_reduced = fit_remove_diameter
summary(fit1_reduced)

plot(fit1_reduced)
```

# A.17 Performance Evaluation of First Order Models

```{r}
#Performance Evaluation: there is not much difference between full and reduced first order models
#Reduced model has lower AIC, BIC
rmspe <- function(y, yh) sqrt(mean((y - yh)^2))

#fit1 with all predictors, transformed response
fit1_rmspe = c(rmspe(train$rings, fit1$fitted.values), rmspe(valid$rings, predict(fit1, valid)))
fit1_aic = n*log(sum((fit1$residuals)^2)/n) + 2*9
fit1_aic # 
fit1_bic = n*log(sum((fit1$residuals)^2)/n) + log(n)*9
fit1_bic #

#fit1_reduced without whole,diameter, with transformed response
fit1_reduced_rmspe = c(rmspe(train$rings, fit1_reduced$fitted.values), rmspe(valid$rings, predict(fit1_reduced, valid)))
fit1_reduced_aic = n*log(sum((fit1_reduced$residuals)^2)/n) + 2*7
fit1_reduced_aic # -
fit1_reduced_bic = n*log(sum((fit1_reduced$residuals)^2)/n) + log(n)*7
fit1_reduced_bic #-
```


# A.18 First Order Model with Sex Reduced to Two Classes

```{r}
#Fit a first order model on training set with consolidated M,F classes
fit1_infant = lm(rings~.-sex,data=train)
summary(fit1_infant)
plot(fit1_infant)
```

# A.19 Reduced First Order Model with Sex Reduced to Two Classes

```{r}
#Fit a reduced first order model on training set with consolidated M,F classes
fit1_infant_reduced = lm(rings~.-sex-diameter-whole,data=train)
summary(fit1_infant_reduced)
plot(fit1_infant_reduced)
```

# A.20 Performance Evaluation of A.18 and A.19

```{r}
#fit1 infant with all predictors, transformed response
fit1_infant_rmspe = c(rmspe(train$rings, fit1$fitted.values), rmspe(valid$rings, predict(fit1, valid)))
fit1_infant_aic = n*log(sum((fit1_infant$residuals)^2)/n) + 2*9
fit1_infant_aic # 
fit1_infant_bic = n*log(sum((fit1_infant$residuals)^2)/n) + log(n)*9
fit1_infant_bic #

#fit1_reduced infant without whole,diameter, with transformed response
fit1_infant_reduced_rmspe = c(rmspe(train$rings, fit1_reduced$fitted.values), rmspe(valid$rings, predict(fit1_reduced, valid)))
fit1_infant_reduced_aic = n*log(sum((fit1_infant_reduced$residuals)^2)/n) + 2*7
fit1_infant_reduced_aic # 
fit1_infant_reduced_bic = n*log(sum((fit1_infant_reduced$residuals)^2)/n) + log(n)*7
fit1_infant_reduced_bic # 
```


# A.21 Summary of Performance Evaluation for First Order Models

```{r}
fit1_eval = cbind(summary(fit1)$adj.r.squared,fit1_aic,fit1_bic,fit1_rmspe[1],fit1_rmspe[2])
fit1_reduced_eval = cbind(summary(fit1_reduced)$adj.r.squared,fit1_reduced_aic,fit1_reduced_bic,fit1_reduced_rmspe[1],fit1_reduced_rmspe[2])
fit1_infant_eval = cbind(summary(fit1_infant)$adj.r.squared,fit1_infant_aic,fit1_infant_bic,fit1_infant_rmspe[1],fit1_infant_rmspe[2])
fit1_infant_reduced_eval = cbind(summary(fit1_infant_reduced)$adj.r.squared,fit1_infant_reduced_aic,fit1_infant_reduced_bic,fit1_infant_reduced_rmspe[1],fit1_infant_reduced_rmspe[2])

eval = rbind(fit1_eval,fit1_reduced_eval,fit1_infant_eval,fit1_infant_reduced_eval)
colnames(eval) = c('R^2_a','AIC','BIC','Training RMSPE','Testing RMSPE')


rownames(eval) = c('First Order - Full','First Order - Reduced','First Order - Full, 2 Class Sex','First Order - Reduced, 2 Class Sex')

eval
```



# A.22 Centering Predictor Variables

```{r}
#Note there appears to be a linear relationship between age vs length,diameter,height; a quadratic relationship with whole, shucked,viscera,shell
#Fit an appropriate corresponding second order model on training set
#First center the quantitative predictor variables
center_train = cbind(train$sex,train$length-mean(train$length),train$diameter-mean(train$diameter),train$height-mean(train$height),train$whole-mean(train$whole),train$shucked-mean(train$shucked),train$viscera-mean(train$viscera),train$shell-mean(train$shell),train$rings-mean(train$rings),train$infant)
center_train = data.frame(center_train)
colnames(center_train) = colnames(train)
center_train$sex = as.factor(center_train$sex)
center_train$infant = as.factor(center_train$infant)

center_valid = cbind(valid$sex,valid$length-mean(valid$length),valid$diameter-mean(valid$diameter),valid$height-mean(valid$height),valid$whole-mean(valid$whole),valid$shucked-mean(valid$shucked),valid$viscera-mean(valid$viscera),valid$shell-mean(valid$shell),valid$rings-mean(valid$rings),valid$infant)
center_valid = data.frame(center_valid)
colnames(center_valid) = colnames(valid)
center_valid$sex = as.factor(center_valid$sex)
center_valid$infant = as.factor(center_valid$infant)
```

# A.23 Second Order Model with Select Quadratic Terms

```{r}
fit2 = lm(rings~.-infant+I(whole^2)+I(shucked^2)+I(viscera^2)+I(shell^2),data=center_train)
summary(fit2)
plot(fit2)
```

# A.24 Second Order Model with All Quadratic Terms

```{r}
#Fit the full second order model for comparison
fit2_full = lm(rings~.-infant+I(length^2)+I(diameter^2)+I(height^2)+I(whole^2)+I(shucked^2)+I(viscera^2)+I(shell^2),data=center_train)
summary(fit2_full)
plot(fit2_full)
```

# A.25 Performance Evaluation of A.23

```{r}
#Performance Evaluation of second order model
fit2_rmspe = c(rmspe(center_train$rings, fit2$fitted.values), rmspe(center_valid$rings, predict(fit2, center_valid)))
fit2_aic = n*log(sum((fit2$residuals)^2)/n) + 2*13
fit2_aic # -15348.83
fit2_bic = n*log(sum((fit2$residuals)^2)/n) + log(n)*13
fit2_bic #-15247.44

fit2_eval = cbind(summary(fit2)$adj.r.squared,fit2_aic,fit2_bic,fit2_rmspe[1],fit2_rmspe[2])
```

# A.26 Performance Evaluation of A.24

```{r}
#Performance Evaluation of second order model full size
fit2_full_rmspe = c(rmspe(center_train$rings, fit2_full$fitted.values), rmspe(center_valid$rings, predict(fit2_full, center_valid)))
fit2_full_aic = n*log(sum((fit2_full$residuals)^2)/n) + 2*16
fit2_full_aic # -15348.83
fit2_full_bic = n*log(sum((fit2_full$residuals)^2)/n) + log(n)*16
fit2_full_bic #-15247.44

fit2_full_eval = cbind(summary(fit2_full)$adj.r.squared,fit2_full_aic,fit2_full_bic,fit2_full_rmspe[1],fit2_full_rmspe[2])

```


# A.27 Summary of Performance Evaluations for Second Order Models

```{r}
eval2 = rbind(fit2_eval,fit2_full_eval)
colnames(eval2) = c('R^2_a','AIC','BIC','Training RMSPE','Testing RMSPE')


rownames(eval2) = c('Second Order - Select Quadratic Terms','Second Order - All Quadratic Terms')

eval2
```


# A.28 Stepwise Regression on A.24

```{r}
#Fits using stepwise regression
library(leaps)

empty = lm(rings~1,data=center_train)

forward_aic = stepAIC(empty,scope=list(upper=fit2_full,lower=empty),direction='forward',k=2,trace=0)
forward_bic = stepAIC(empty,scope=list(upper=fit2_full,lower=empty),direction='forward',k=log(n),trace=0)

backward_aic = stepAIC(fit2_full,scope=list(upper=fit2_full,lower=empty),direction='backward',k=2,trace=0)
backward_bic = stepAIC(fit2_full,scope=list(upper=fit2_full,lower=empty),direction='backward',k=log(n),trace=0)

both_aic = stepAIC(empty,scope=list(upper=fit2_full,lower=empty),direction='both',k=2,trace=0)
both_bic = stepAIC(empty,scope=list(upper=fit2_full,lower=empty),direction='both',k=log(n),trace=0)
```

# A.29 Stepwise Regression Second Order Models 

```{r}
#Forward stepwise models
#summary(forward_aic)
#forward_aic$anova

#summary(forward_bic)
#forward_bic$anova

anova(forward_aic,forward_bic)
```



```{r}
#Backward stepwise models
#summary(backward_aic)
#backward_aic$anova

#summary(backward_bic)
#backward_bic$anova

anova(backward_aic,backward_bic)
```


```{r}
#Bidirectional stepwise models
#summary(both_aic)
#both_aic$anova

#summary(both_bic)
#both_bic$anova

anova(both_aic,both_bic)
```

# A.30 Performance Evaluation of A.29 Bidirectional Stepwise Regression Second Order Model Using BIC

```{r}
#Performance Evalaution for birectional stepwise bic on fit2 full
fit2_full_stepBIC_RMSPE = c(rmspe(center_train$rings, both_bic$fitted.values), rmspe(center_valid$rings, predict(both_bic, center_valid)))
fit2_full_stepBIC_aic = n*log(sum((both_bic$residuals)^2)/n) + 2*10
#fit2_full_aic # -15348.83
fit2_full_stepBIC_bic = n*log(sum((both_bic$residuals)^2)/n) + log(n)*10
#fit2_full_bic #-15247.44

fit2_full_stepBIC_eval = cbind(summary(both_bic)$adj.r.squared,fit2_full_stepBIC_aic,fit2_full_stepBIC_bic,fit2_full_stepBIC_RMSPE[1],fit2_full_stepBIC_RMSPE[2])
#fit2_full_stepBIC_eval
eval2 = rbind(eval2,fit2_full_stepBIC_eval)

rownames(eval2) = c('Second Order - Select Quadratic Terms','Second Order - All Quadratic Terms','Stepwise Regression (BIC) on Second Order')

eval2
```


# A.31 Summary of Final Model

```{r}
summary(both_bic)
```
# A.32 Final Model Diagnostics

```{r}
plot(both_bic)
```